---
title: "p8105_hw3_pm2995"
author: "Priyal"
date: "10/8/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)
library(readxl)
library(p8105.datasets)
library(ggplot2)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


#Problem 1

##Problem 1.1

Loading the data and some cleaning 

First we load the BRFSS data from the `p8105.datasets` package.

```{r}
data(brfss_smart2010)
```

Next, we do some cleaning to ensure appropriate variable names. I cleaned the data using janitor:: clean_names function converting them all to lower snake case, filter to include only overall health topic data and include responses from excellent to poor. Further, I organized the responses as a factor taking levels ordered from excellent to poor using mutate.

```{r tidy_brfss}
brfss_tidy = brfss_smart2010 %>% 
  janitor::clean_names() %>%
  filter(topic == "Overall Health") %>%
  filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) %>% 
  rename(location_abbr = locationabbr, location_desc = locationdesc) %>%
  mutate(response = as.factor(response)) %>% 
  mutate(
    response = fct_relevel(response, c("Excellent", "Very good", "Good", "Fair", "Poor"))
    )
```

