---
title: "p8105_hw3_pm2995"
author: "Priyal"
date: "10/8/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)
library(readxl)
library(p8105.datasets)
library(ggplot2)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


#Problem 1

##Problem 1.1

Loading the data and some cleaning 

First we load the BRFSS data from the `p8105.datasets` package.

```{r}
data(brfss_smart2010)
```

Next, we do some cleaning to ensure appropriate variable names. I cleaned the data using janitor:: clean_names function converting them all to lower snake case, filter to include only overall health topic data and include responses from excellent to poor. Further, I organized the responses as a factor taking levels ordered from excellent to poor using mutate.

```{r tidy_brfss}
brfss_tidy = brfss_smart2010 %>% 
  janitor::clean_names() %>%
  filter(topic == "Overall Health") %>%
  filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) %>% 
  rename(location_abbr = locationabbr, location_desc = locationdesc) %>%
  mutate(response = as.factor(response)) %>% 
  mutate(
    response = fct_relevel(response, c("Excellent", "Very good", "Good", "Fair", "Poor"))
    )
```


Specific Questions

##Problem 1.2

In 2002, which states were observed at 7 locations?

Answer:
The code chunk below filters to include the rows from year 2002 and groups the data by location_abbr (which includes states). Further, using summarize number of distinct locations in each state has been taken out. I have used filter to keep states that were observed at 7 locations.

```{r}
brfss_tidy %>% 
  filter(year == 2002) %>% 
  group_by(location_abbr) %>% 
  summarize(n_locations = n_distinct(location_desc)) %>% 
  filter(n_locations == 7)
```

Three states were observed at seven locations: CT, FL, and NC.

##Problem 1.3

Make a “spaghetti plot” that shows the number of locations in each state from 2002 to 2010.

Answer:
The code below groups the brfss_tidy data by location_abbr(state) and year and further uses summarize to take out the number of corresponding observations. It then creates a spaghetti plot that shows the number of locations in each state from 2002 to 2010. 


```{r}
brfss_tidy %>% 
  group_by(location_abbr, year) %>% 
  summarize(n_obs = n()) %>%
  ggplot(aes(x = year, y = n_obs, color = location_abbr)) +
  geom_line() + labs(title = "No. of locations in each state from 2002 to 2010", x = "Year", y = "Number of observations")
```
